# Setup R Packages {-}

Some setup is required to reproduce this book. Here are key R packages you should consider installing and specific instructions for Windows and Linux users.

## R Packages {-}

In this book I discuss how to use a number of user-written R packages for reproducible research. Many of these packages are not included in the default R installation. They need to be installed separately.

\index{R!packages|(}

**Note:** in general you should aim to minimize the number of packages that your research depends on. Doing so will lessen the possibility that your code will ``break'' when a package is updated. This book depends on relatively many packages because of its special and unusual purpose of illustrating a variety of tools that you can use for reproducible research.

To install key user-written packages discussed in this book, copy the following code and paste it into your R console:

```{r package_install, results='hide'}
# Packages to install
pkg_to_install <- c("brew", "bookdown", "xfun")


# Check if the packages are installed, if not install
lapply(pkg_to_install,
    function(pkg) {
  if (system.file(package = pkg) == '') install.packages(pkg)
})
```

## Install LaTeX

\index{R!packages|(}

Once you enter this code, you may be asked to select a CRAN "mirror"\index{CRAN!mirror} to download the packages from.^[CRAN stands for the Comprehensive R Archive Network.] Simply select the mirror closest to you.

## Special issues for Windows and Linux Users {-}

\index{Windows|(}

If you are using Windows, you will also need to install *Rtools* \cite[]{Rtools}.\index{Rtools} You can install *Rtools* from: \url{http://cran.r-project.org/bin/windows/Rtools/}.\label{RtoolsDownload} Please use the recommended installation to ensure that your system PATH\index{PATH} is set up correctly. Otherwise your computer will not know where the tools are. Alternatively, use the `install.Rtools()` function from the *installr* [@galili2018]\index{installr)} package to install it.

\index{Windows|(}

\index{Linux|(}

On Linux you will need to install the \emph{RCurl} \citep{R-RCurl}\index{RCurl} and \emph{XML}\index{XML} \citep{R-XML} packages separately. Use your Terminal\index{Terminal} to install these packages with the following code:

```{sh, eval=FALSE}
sudo apt-get update

sudo apt-get install libcurl4-gnutls-dev
sudo apt-get install libxml2-dev
sudo apt-get install r-cran-xml
sudo apt-get install r-cran-rjava
```

\index{Linux|(}

```{r sessioninfo}
xfun::session_info()
```
