# Getting Started with Reproducible Research {#GettingStartedRR}

Researchers often start thinking about making their work reproducible
near the end of the research process when they write up their results or
maybe even later when a journal requires their data and code be made
available for publication. Or maybe even later when another researcher
asks if they can use the data from a published article to reproduce the
findings. By then there may be numerous versions of the data set and
records of the analyses stored across multiple folders on the
researcher's computers. It can be difficult and time consuming to sift
through these files to create an accurate account of how the results
were reached. Waiting until near the end of the research process to
start thinking about reproducibility can lead to incomplete
documentation that does not give an accurate account of how findings
were made. Focusing on reproducibility from the beginning of the process
and continuing to follow a few simple guidelines throughout your
research can help you avoid these problems. Remember "reproducibility is
not an afterthought–it is something that must be built-into the project
from the beginning" [@donoho2010 386].

This chapter first gives you a brief overview of the reproducible
research process: a workflow for reproducible research. Then it covers
some of the key guidelines that can help make your research more
reproducible.

## The Big Picture: A Workflow for Reproducible Research

The three basic stages of a typical computational empirical research
project are:

-   data gathering,

-   data analysis,

-   results presentation.

Each stage is part of the reproducible research workflow covered in this
book. Tools for reproducibly gathering data are covered in Part II. Part
III teaches tools for tying the data we gathered to our statistical
analyses and presenting the results with tables and figures. Part IV
discusses how to tie these findings into a variety of documents you can
use to advertise your findings.

Instead of starting to use the individual tools of reproducible research
as soon as you learn them, I recommend briefly stepping back and
considering how the stages of reproducible research *tie* together.
This will make your workflow more coherent from the beginning
and save you a lot of backtracking later on. Figure \@ref(fig:WorkflowTies)
illustrates the workflow. Notice that most of the arrows connecting the
workflow's parts point in both directions, indicating that you should
always be thinking about how to make it easier to go backwards through
your research, i.e. reproduce it, as well as forwards.

Around the edges of the figure are some of the functions you will learn
to make it easier to go forwards and backwards through the process.
These functions tie your research together. For example, you can use
API-based R packages to gather data from the internet. You can use R's
`merge()`\index{R function!merge} function to combine data gathered from different sources into one
data set. The `getURL()`\index{R function!getURL} function from R's *RCurl* package [@R-RCurl] and
the `read.table()`\index{R function!read.table} function in base R or the much more versatile `import()` 
function from the *rio* package [@R-rio]\index{rio}\index{R function!import} can be used to bring this data set into your statistical analyses. The *knitr* or *rmarkdown* package then
ties your analyses into your presentation documents. This includes the
code you used, the figures you created, and, with the help of tools such
as the `kable()` function in the *knitr* package, tables of results. You
can even tie multiple presentation documents together. For example, you
can access the same figure for use in a LaTeX article and a
Markdown-created website with the `includegraphics`\index{LaTeX command!include\_graphics} and *knitr*'s `include_graphics()`\index{R function!include\_graphics}
functions, respectively. This helps you maintain a consistent
presentation of results across multiple document types. We'll cover
these functions in detail throughout the book. See Table
\@ref(TableTieFunctions) for an additional overview of some of the
*tie functions*.

```{r WorkflowTies, engine = "tikz", fig.cap = "Example Workflow and a Selection of Functions to Tie It Together", cache=TRUE, echo=FALSE, fig.ext=if (knitr:::is_latex_output()) 'pdf' else 'png'}
\usetikzlibrary{decorations.pathmorphing}

\definecolor{Blue}{HTML}{7BCCC4}
\definecolor{LiteBlue}{HTML}{A8DDB5}
\definecolor{DarkBlue}{HTML}{08589E}

\definecolor{GrayLine}{HTML}{BDBDBD}

% Set node styles
%% Workflow stage nodes
\tikzstyle{Stage} = [draw=Blue,
                     %fill=Blue,
                     rectangle,
                     text width=7em,
                     inner sep=0.5cm,
                     font=\small]

% Raw Data nodes
\tikzstyle{RawData} = [draw=LiteBlue,
                       %fill=LiteBlue,
                       decorate,
                       decoration={random steps,
                                   segment length=2pt,
                                   amplitude=2pt},
                       inner sep=0.25cm,
                       font=\scriptsize]

% Separator line style
\tikzstyle{sepline} = [draw,
                        very thick,
                        color=GrayLine]

% Link function nodes
\tikzstyle{Links} = [draw=none,
                          text width=6em,
                          text=DarkBlue,
                          font=\footnotesize]

% Begin tikz picture
\begin{tikzpicture}

    % Raw Data Nodes
    \node (Data1) at (-3, 7) [RawData]{Raw Data};
    \node (Data2) at (-3, 5) [RawData]{Raw Data};
    \node (Data3) at (-3, 3) [RawData]{Raw Data};

    % Workflow stage nodes
    \node (DataGather) at (0.5, 5) [Stage, text width= 6em]{Data Gather};
    \node (Analysis) at (5.5, 5) [Stage, text width= 4em]{Analysis};
    \node (Presentation1) at (9, 8) [Stage]{LaTeX Book, \\ Article, \& \\ Slideshow \\ Presentations};
    \node (Presentation2) at (9, 2.5) [Stage]{Markdown/ \\ HTML Website \\ Presentations};

    % Lines
    \draw [->, very thick] (Data1) -- (DataGather);
    \draw [->, very thick] (Data2) -- (DataGather);
    \draw [->, very thick] (Data3) -- (DataGather);
    \draw [<->, very thick] (DataGather) -- (Analysis);
    \draw [<->, very thick] (Analysis) -- (Presentation1);
    \draw [<->, very thick] (Analysis) -- (Presentation2);

    \draw [<->, very thick] (Presentation1) -- (Presentation2);

    \path [sepline] (-3.5, 0.75) -- (11, 0.75);
    \path [sepline] (11.5, 9) -- (11.5, 1.5);

    % Link function nodes

    \node (pres) at (13, 5) [Links]{{\emph{knitr}} \\ \texttt{input} \\ \texttt{include} \\ \texttt{includegraphics} \\ \texttt{include\_graphics} \\ Pandoc \\ \texttt{![]()}};
    \node (knitr) at (7.5, -1) [Links]{ {\emph{knitr}} \\ \emph{rmarkdown} \\ \texttt{source} \\ \texttt{source\_url} \\ \texttt{kable} \\ \texttt{print(xtable())} \\ \texttt{texreg} };
    \node (readData) at (3, -1) [Links]{\texttt{import} \\ \texttt{read.table} \\ \texttt{getURL} };

    \node (importData) at (-1, -1.3) [Links]{ \texttt{Make} \\ \texttt{download.file} \\ \texttt{read.table} \\ \texttt{import} \\ \texttt{merge}\\ \texttt{getURL} \\ API-based packages };


\end{tikzpicture}
```

### Reproducible theory

An important part of the research process that I do not discuss in this
book is theoretical stage. Ideally, if you are using a deductive
research design, the bulk of this work will precede and guide the data
gathering and analysis stages. Just because I don't cover this stage of
the research process doesn't mean that theory building can't and
shouldn't be reproducible. It can in fact be "the easiest part to make
reproducible" [@vandewalle2007 1254]. Quotes and paraphrases from
previous works in the literature obviously need to be fully cited so
that others can verify that they accurately reflect the source material.
For mathematically based theory, you should give clear and complete descriptions of the
proofs.

Though I don't actively cover theory replication in depth in this book,
I do touch on some of the ways to incorporate proofs and citations into
your presentation documents. These tools are covered in Part IV.

## Practical Tips for Reproducible Research

Before we start learning the details of the reproducible research
workflow with R and RStudio, it's useful to cover a few broad tips that
will help you organize your research process and put these skills in
perspective. The tips are:

1.  Document everything!

2.  Everything is a (text) file.

3.  All files should be human readable.

4.  Explicitly tie your files together.

5.  Have a plan to organize, store, and make your files available.

Using these tips will help make your computational research really
reproducible.

### Document everything!

In order to reproduce your research, others must be able to know what
you did. You have to tell them what you did by documenting as much of
your research process as possible. Ideally, you should tell your readers
how you gathered your data, analyzed it, and presented the results.
Documenting everything is the key to reproducible research and lies
behind all of the other tips in this chapter and tools you will learn
throughout the book.

#### Document your R session info {- #SessionInfoHow}

Before discussing the other tips it's important to learn a key part of
documenting with R. You should *record your session info*. Many things
in R have stayed the same since it was introduced in the early 1990s.
This makes it easy for future researchers to recreate what was done in
the past. However, things can change from one version of R to another
and especially from one version of an R package to another. Also, the
way R functions and how R packages are handled can vary across different
operating systems, so it's important to note what system you used.
Finally, you may have R set to load packages by default (see Section
\@ref(Packages) for information about packages). These packages might be
necessary to run your code, but other people might not know what
packages and what versions of the packages were loaded from just looking
at your source code. The `sessionInfo()` function in R prints a record of
all of these things. The information from the session I used to create
this book is:

```{r Ch2SessionInfoPlain, echo=TRUE, eval=FALSE}
# Print R session info
sessionInfo()
```

```{r RemoveLongExtra, echo=FALSE}
si <- sessionInfo()

# Remove non-relevant information that doesn't fit on the page
si$BLAS <- NULL
si$LAPACK <- NULL

si
```

Chapter \@ref(DirectoriesChapter) gives specific details about how to
create files with dynamically included session information. If you use
non-R tools you should also record what versions of these tools you
used.

### Everything is a (text) file

Your documentation is stored in files that include data, analysis code,
the write-up of results, and explanations of these files (e.g. data set
codebooks, session info files, and so on). Ideally, you should use the
simplest file format possible to store this information. Usually the
simplest file format is the humble, but versatile, text file.[^chapter2_1]

Text files are extremely nimble. They can hold your data in, for
example, comma-separated values (CSV) format. They can contain your
analysis code in files. And they can be the basis for your presentations
as markup documents like or , for LaTeX and Markdown files,
respectively. All of these files can be opened by any program that can
read text files.

One reason reproducible research is best stored in text files is that
this helps *future-proof* your research. Other file formats, like
those used by Microsoft Word (`.docx`) or Excel (`.xlsx`), change
regularly and may not be compatible with future versions of these
programs. Text files, on the other hand, can be opened by a very wide
range of currently existing programs and, more likely than not, future
ones as well. Even if future researchers do not have R or a LaTeX
distribution, they will still be able to open your text files and, aided
by frequent comments (see below), be able to understand how you
conducted your research [@bowers2011 3].

Text files are also very easy to search and manipulate with a wide range
of programs–such as R and RStudio–that can find and replace text
characters as well as merge and separate files. Finally, text files are
easy to version and changes can be tracked using programs such as Git
(see Chapter \@ref(Storing)).

#### Learn from the text file: keep it simple {-}

Text files are simple. Their simplicitly increases the probability of baseline usefulness in the future to researchers who will reproduce the work. We can extend the logic of the simple text file to all of the tools we use: keep it simple. Avoid adding dependencies you don't need to actually gather your data, analyze it, and present the results. For example, I have been tempted to make my presentation slides look nicer with custom fonts. I was later burned when I wanted to make minor changes to slides a year after I first presented them (and a day before teaching an upcoming class) only to find that the custom fonts were no longer available. This broke my slides and forced me to spend considerable time reworking writing my source documents. If I, the creator of the slides, found this time consuming and annoying, an independent researcher would likely find it even more difficult.

### All files should be human readable

Treat all of your research files as if someone who has not worked on the
project will, in the future, try to understand them. Computer code is a
way of communicating with the computer. It is ‘machine readable' in that
the computer is able to use it to understand what you want to do.[^chapter2_2]
However, there is a very good chance that other people (or you six
months in the future) will not understand what you were telling the
computer. So, you need to make all of your files ‘human readable'. To
make them human readable, you should comment on your code with the goal
of communicating its design and purpose [@wilson2012]. With this in mind
it is a good idea to *comment frequently* [@bowers2011 3] and
*format your code using a style guide* [@nagler1995]. For especially
important pieces of code you should use *literate programming*–where
the source code and the presentation text describing its design and
purpose appear in the same document.\index{literate programming} Doing this will make it very clear to others how you accomplished a piece of research.

#### Commenting {-}

In R, everything on a line after a hash character––(also known as
'number', 'pound', or 'sharp') is ignored by R, but is readable to people who
open the file. The hash character is a comment declaration character.
You can use the to place comments telling other people what you are
doing. Here are some examples:

```{r Ch2CommentHash}
# A complete comment line
2 + 2 # A comment after R code
```

On the first line the hash is placed at the very beginning, so the
entire line is treated as a comment. On the second line the is placed
after the simple equation `2 + 2`. R runs the function and finds the
answer `4`, but it ignores all of the words after the hash.

Different languages have different comment declaration characters. In
LaTeX everything after the percent sign is treated as a comment, and in
Markdown/HTML comments are placed inside of `<!- ->`. The hash character is used
for comment declaration in command-line shell scripts as well as many other programming languages such as Python and Julia.\index{Python}\index{Julia}

Nagler [-@nagler1995 491] gives some advice on when and how to use
comments:

-   write a comment before a block of code describing what the code
    does,

-   comment on any line of code that is ambiguous.

In this book I follow these guidelines when displaying code. Nagler also
suggests that all of your source code files should begin with a comment
header. At the least, the header should include:

-   a description of what the file does,

-   the date it was last updated,

-   the name of the file's creator and any contributors.

You may also want to include other information in the header such as
what files it depends on, what output files it produces, what version of
the programming language you are using, sources that may have influenced
the code, and how the code is licensed. Here is an example of a minimal
file header for an R source code file that creates the third figure in
an article titled ‘My Article':

````R
############################
# R Source code file used to create Figure 3 in 'My Article'
# Created by Christopher Gandrud
# MIT License
############################
````

Feel free to use things like the long series of hash marks above and
below the header, white space, and indentations to make your comments
more readable.

#### Style guides {-}

In natural language writing you don't necessarily have to follow a style
guide. People could probably figure out what you are trying to say, but
it is a lot easier for your readers if you use consistent rules. The
same is true when writing computer code. It's good to follow consistent
rules for formatting your code so that it's easier for you and others to
understand.

There are a number of R style guides. Most of them are similar to the
Google R Style Guide.[^chapter2_3] Hadley Wickham also has a nicely presented R
style guide.[^chapter2_4] You may want to use the *styler* [@R-styler]\index{styler}
package to automatically reformat your code so that it is easier to
read.

#### Literate programming {-}

For particularly important pieces of research code it may be useful to
not only comment on the source file, but also display code in
presentation text. For example, you may want to include key parts of the
code you used for your main statistical models and an explanation of
this code in an appendix following your article. This is commonly
referred to as literate programming [@knuth1992].\index{literate programming}

### Explicitly tie your files together

If everything is just a text file, then research projects can be thought
of as individual text files that have a relationship with one another.
They are tied together. A data file is used as input for an analysis
file. The results of an analysis are shown and discussed in a markup
file that is used to create a PDF document. Researchers often do not
explicitly document the relationships between files that they used in
their research. For example, the results of an analysis–a table or
figure–may be copied and pasted into a presentation document. It can be
very difficult for future researchers to trace the table or figure back
to a particular statistical model and a particular data set without
clear documentation. Therefore, it is important to make the links
between your files explicit.

Tie functions are the most dynamic way to explicitly link your files
together. These functions instruct the computer program you are using to
use information from another file. In Table \@ref(TableTieFunctions) I have
compiled a selection of key tie functions you will learn how to use in
this book. We'll discuss many more, but these are some of the most
important.

### Have a plan to organize, store, and make your files available

Finally, in order for independent researchers to reproduce your work,
they need to be able access the files that instruct them how to do this.
Files also need to be organized so that independent researchers can
figure out how they fit together. So, from the beginning of your
research process you should have a plan for organizing your files and a
way to make them accessible.

One rule of thumb for organizing your research in files is to limit the
amount of content any one file has. Files that contain many different
operations can be very difficult to navigate, even if they have detailed
comments. For example, it would be very difficult to find any particular
operation in a file that contained the code used to gather the data, run
all of the statistical models, and create the results figures and
tables. If you have a hard time finding things in a file you created,
think of the difficulties independent researchers will have!

Because we have so many ways to link files together, there is really no
need to lump many different operations into one file. So, we can make
our files modular. One source code file should be used to complete one
or just a few tasks. Breaking your operations into discrete parts will
also make it easier for you and others to find errors [@nagler1995 490].

Chapter \@ref(DirectoriesChapter) discusses file organization in much more
detail. Chapter \@ref(Storing) teaches you a number of ways to make your
files accessible through the cloud computing services like GitHub.

\begin{table}
    \caption{A Selection of Functions/Packages/Programs for Tying Together Your Research Files}
    \label{TableTieFunctions}
    \vspace{0.3cm}
    {\footnotesize{
    \begin{tabular}{p{2.5cm} c p{5.25cm} p{2cm}}
        \hline
        Function/Package/ Program & Language & Description & Chapters Discussed \\[0.3cm]
        \hline \hline
        {\emph{knitr}} & R & R package with commands for tying analysis code into presentation documents including those written in LaTeX and Markdown. & \hfill Throughout \\[0.25cm]
        \emph{rmarkdown} & R & R package that builds on \emph{knitr}. It allows you to use Markdown to output to HTML, PDFs compiled with LaTeX or Microsoft Word. & \hfill Throughout \\[0.25cm]
        {\tt{download.file}} & R & Downloads a file from the internet. & \hfill\ref{DataGather} \\[0.25cm]
        {\tt{read.table}} & R & Reads a table into R. You can use this to import a plain-text file formatted data into R. & \hfill\ref{DataGather} \\[0.25cm]
        {\tt{read.csv}} & R & Same as \texttt{read.table} with default arguments set to import \texttt{.csv} formatted data files. & \hfill\ref{DataGather} \\[0.25cm]
        {\tt{import}} & R & Reads a table stored locally or on the internet into R. You can use it to import a wide variety of plain-text data formats into R from secure (https) URLs. & \hfill\ref{DataGather} \\[0.25cm]
        API-based packages & R & Various packages use APIs to gather data from the internet. & \hfill\ref{DataGather} \\[0.25cm]
        {\tt{merge}} & R & Merges together data frames. & \hfill\ref{DataClean} \\[0.25cm]
        {\tt{source}} & R & Runs an R source code file. & \hfill\ref{StatsModel} \\[0.25cm]
        {\tt{source\_url}} & R & From the {\emph{devtools}} package. Runs an R source code file from a secure ({\tt{https}}) url like those used by GitHub. & \hfill\ref{StatsModel} \\[0.25cm]
        {\tt{kable}} & R & Creates tables from data frames that can be rendered using Markdown or LaTeX. & \hfill\ref{TablesChapter} \\[0.25cm]
        {\tt{toLaTeX}} & R & Converts R objects to LaTeX. & \hfill\ref{GettingStartedRR} \\[0.25cm]
        {\tt{input}} & LaTeX & Includes LaTeX files inside of other LaTeX files. & \hfill\ref{LargeDocs} \\[0.25cm]
        {\tt{include}} & LaTeX & Similar to {\tt{input}}, but puts page breaks on either side of the included text. Usually it is used for including chapters. & \hfill\ref{LargeDocs} \\[0.25cm]
        {\tt{includegraphics}} & LaTeX & Inserts a figure into a LaTeX document. & \hfill\ref{FiguresChapter} \\[0.25cm]
        {\tt{include\_graphics}} & R/R Markdown & Inserts a figure into an R Markdown document. & \hfill\ref{FiguresChapter} \\[0.25cm]
        \texttt{![]()} & Markdown & Inserts a figure into a Markdown document. & \hfill\ref{MarkdownChapter} \\  [0.25cm]
        Pandoc & shell & A shell program for converting files from one markup language to another. Allows you to tie presentation documents together. & \hfill\ref{LargeDocs} \& \ref{MarkdownChapter} \\[0.25cm]
        Make & shell & A shell program for automatically building many files at the same time. & \hfill\ref{DataGather} \\[0.25cm]
        \hline
    \end{tabular}
    }}
\end{table}

[^chapter2_1]: Plain text files are usually given the file extension `.txt`.
    Depending on the size of your data set it may not be feasible to
    store it as a text file. Nonetheless, text files can still be used
    for analysis code and presentation files.

[^chapter2_2]: Of course, if the computer does not understand it will usually
    give an error message.

[^chapter2_3]: See:
    <http://google-styleguide.googlecode.com/svn/trunk/google-r-style.html>.

[^chapter2_4]: You can find it at <http://adv-r.had.co.nz/Style.html>.
